---
title: "A Quick Introduction to qIE via Examples"
author: "Kai-Hsiang Hu and Anne Chao"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    vignette: >
      %\VignetteIndexEntry{A Quick Introduction to qIE via Examples}
      %\VignetteEngine{knitr::rmarkdown}
      %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, 
                      comment = "", 
                      fig.retina = 2,
                      fig.align = 'center',
                      fig.width = 7, 
                      fig.height = 4)
options("width"=200)

library(iNEXT.IE)
```


`iNEXT.IE` (INterpolation and EXTrapolation for Inter-specific Encounter) is a framework for biodiversity methodology. 

`iNEXT.IE` features two statistical analyses (non-asymptotic and asymptotic):

(1) A non-asymptotic approach based on interpolation and extrapolation for diversity 

`iNEXT.IE` computes the estimated diversity for standardized samples with a common sample size or sample completeness. This approach aims to compare diversity estimates for equally-large (with a common sample size) or equally-complete (with a common sample coverage) samples; it is based on the seamless rarefaction and extrapolation (R/E) sampling curves of diversity for q = 0, 1 and 2. `iNEXT.IE` offers three types of R/E sampling curves:


* Sample-size-based (or size-based) R/E sampling curves: This type of sampling curve plots the diversity estimates with respect to sample size.

* Coverage-based R/E sampling curves: This type of sampling curve plots the diversity estimates with respect to sample coverage.

* Sample completeness curve: This curve depicts how sample coverage varies with sample size. The sample completeness curve provides a bridge between the size- and coverage-based R/E sampling curves.

(2) An asymptotic approach to infer asymptotic diversity  

`iNEXT.IE` computes the estimated asymptotic diversity and also plots diversity profiles (q-profiles) for q between 0 and 2, in comparison with the maximum likelihood estimation. Typically, the asymptotic estimates for q $\geq$ 1 are reliable, but for q < 1 (especially for q = 0, species richness), the asymptotic estimates represent only lower bounds. 


## How to cite
If you publish your work based on results from `qIE` package, you should make references to the following methodology paper and the package:

- Hu, K.-H. and Chao, A. (2025). The iNEXT.IE package: INterpolation and EXTrapolation for Inter-specific Encounter. R package available from CRAN.  


## SOFTWARE NEEDED TO RUN iNEXT.IE IN R
- Required: [R](https://cran.r-project.org/)
- Suggested: [RStudio IDE](https://www.rstudio.com/products/RStudio/#Desktop)


## HOW TO RUN iNEXT.IE:
The `iNEXT.IE` package can be downloaded from CRAN or Kai-Hsiang Hu's [iNEXT.IE_github](https://github.com/KaiHsiangHu/iNEXT.IE) using the commands below. For a first-time installation, some additional packages must be installed and loaded; see package manual. 


```{r eval=FALSE}
## install iNEXT.IE package from CRAN
install.packages("iNEXT.IE")  

## or install the latest version from github
install.packages('devtools')
library(devtools)
install_github('KaiHsiangHu/iNEXT.IE')

## import packages
library(iNEXT.IE)
```


There are six main functions in this package: 

Two functions for non-asymptotic analysis with graphical displays:

- **iNEXTIE** computes standardized diversity estimates of order q = 0, 1 and 2 for rarefied and extrapolated samples at specified sample coverage values and sample sizes.

- **ggiNEXTIE** visualizes the output from the function `iNEXTIE`.


Two functions for point estimation and basic data information

- **estimateIE** computes diversity of order q = 0, 1 and 2 with a particular set of user-specified level of sample sizes or sample coverage values.

- **DataInfoIE** provides basic data information based on the observed data.


Two functions for asymptotic analysis with graphical displays:

- **MLEAsyIE** computes maximum likelihood estimation and asymptotic diversity estimates of order q between 0 and 2 (in increments of 0.2) for diversity.

- **ggMLEAsyIE** visualizes the output from the function `MLEAsyIE`.



## <span style="color:red;">DATA INPUT FORMAT</span>

### Species abundance data format

For `iNEXT.IE` package, information on species identity (or any unique identification code) and assemblage affiliation is not required.

Individual-based abundance data: Data can be input as a vector of species abundances (for a single assemblage), a species (in rows) by assemblage (in columns) matrix/data.frame, or a list of species abundance vectors. 

For example, we generate an example for illustration. 


```{r}
set.seed(2025)
data = list("Site_1" = as.numeric(rmultinom(1, 200, 1:30)), 
            "Site_2" = as.numeric(rmultinom(1, 300, 1:40)))
data
```



## <span style="color:red;">MAIN FUNCTION iNEXTIE(): RAREFACTION/EXTRAPOLATION</span>

We first describe the main function `iNEXTIE()` with default arguments: 

```{r eval=FALSE}
iNEXTIE(data, rho, q = c(0, 1, 2), size = NULL, endpoint = NULL, knots = 40, nboot = 50, conf = 0.95)
```


The arguments of this function are briefly described below, and will be explained in more details by illustrative examples in later text. This main function computes standardized diversity estimates of order q = 0, 1 and 2, the sample coverage estimates, and related statistics for K (if `knots = K` in the specified argument) evenly-spaced knots (sample sizes) between size 1 and the `endpoint`, where the endpoint is described below. Each knot represents a particular sample size for which diversity estimates will be calculated. By default, `endpoint` = double the reference sample size or total individuals. For example, if `endpoint = 10`, `knot = 4` is specified, diversity estimates will be computed for a sequence of samples with sizes (1, 4, 7, 10).  


<table style="width:100%;">
<colgroup>
<col width="20%">
<col width="80%">
</colgroup>
<thead>
<tr class="header">
<th align="center">Argument</th>
<th align="left">Description</th>
</tr>
</thead>
<tbody>
<tr class="odd">
<td align="center"><code>data</code></td>
<td align="left">
data can be input as a vector of species abundances (for a single assemblage), matrix/data.frame (species by assemblages), or a list of species abundance vectors. 
</tr>
<tr class="even">
<td align="center"><code>rho</code></td>
<td align="left">the sampling fraction can be input as a vector for each assemblage or a numeric common value for all assemblages.</td>
</tr>
<tr class="odd">
<td align="center"><code>q</code></td>
<td align="left">a numerical vector specifying the diversity orders. Default is <code>c(0, 1, 2)</code>.</td>
</tr>
<tr class="even">
<td align="center"><code>size</code></td>
<td align="left">an integer vector of sample sizes (number of individuals) for which diversity estimates will be computed.
If <code>NULL</code>, then diversity estimates will be computed for those sample sizes determined by the specified/default <code>endpoint</code> and <code>knots</code>.</td>
</tr>
<tr class="odd">
<td align="center"><code>endpoint</code></td>
<td align="left">an integer specifying the sample size that is the <code>endpoint</code> for rarefaction/extrapolation.
If NULL, then <code>endpoint</code> <code>=</code> double the reference sample size or total individuals.</td>
</tr>
<tr class="even">
<td align="center"><code>knots</code></td>
<td align="left">an integer specifying the number of equally-spaced <code>knots</code> (say K, default is 40) between size 1 and the <code>endpoint</code>; each knot represents a particular sample size for which diversity estimate will be calculated.
If the <code>endpoint</code> is smaller than the reference sample size, then <code>iNEXTIE()</code> computes only the rarefaction estimates for approximately K evenly spaced <code>knots</code>. 
If the <code>endpoint</code> is larger than the reference sample size, then <code>iNEXTIE()</code> computes rarefaction estimates for approximately K/2 evenly spaced <code>knots</code> between sample size 1 and the reference sample size, and computes extrapolation estimates for approximately K/2 evenly spaced <code>knots</code> between the reference sample size and the <code>endpoint</code>.</td>
</tr>
<tr class="odd">
<td align="center"><code>nboot</code></td>
<td align="left">a positive integer specifying the number of bootstrap replications when assessing sampling uncertainty and constructing confidence intervals. Enter 0 to skip the bootstrap procedures. Default is 50.</td>
</tr>
<tr class="even">
<td align="center"><code>conf</code></td>
<td align="left">a positive number < 1 specifying the level of confidence interval. Default is 0.95.</td>
</tr>


</tbody>
</table>


The main function `iNEXTIE()` returns the `iNEXTIE` object, which can be further used to make plots using the function `ggiNEXTIE()` to be described below. The `"iNEXTIE"` object includes three lists: 

(1) `$DataInfo` for summarizing data information. 

(2) `$iNextEst` for showing diversity estimates along with related statistics for a series of rarefied and extrapolated samples; there are two data frames (`$size_based` and `$coverage_based`) conditioning on standardized sample size or sample coverage, respectively.

(3) `$AsyEst` for showing maximum likelihood estimation and asymptotic diversity estimates along with related statistics.

## <span style="color:red;">FUNCTION ggiNEXTIE(): GRAPHIC DISPLAYS</span>

The function `ggiNEXTIE()`, which extends `ggplot2` with default arguments, is described as follows: 

```{r eval=FALSE}
ggiNEXTIE(output, type = 1:3)  
```

Here `output` is the `iNEXTIE()` object. Three types of curves are allowed for diversity:  

(1) Sample-size-based R/E curve (`type = 1`): This curve plots diversity estimates with confidence intervals as a function of sample size.  

(2) Sample completeness curve (`type = 2`): This curve plots the sample coverage with respect to sample size. 

(3) Coverage-based R/E curve (`type = 3`): This curve plots the diversity estimates with confidence intervals as a function of sample coverage. 

The `ggiNEXTIE()` function is a wrapper with the package `ggplot2` to create a rarefaction/extrapolation sampling curve in a single line of code. The figure object is of class `"ggplot"`, so it can be manipulated by using the `ggplot2` tools. 


## <span style="color:blue;">RAREFACTION/EXTRAPOLATION VIA EXAMPLES</span>

### EXAMPLE: rarefaction/extrapolation for abundance data 


```{r, echo=FALSE, cache = TRUE}
output_iNEXT <- iNEXTIE(data, rho = 0.3, q = c(0, 1, 2))
```

Based on the dataset, the following commands return all numerical results. The first list of the output (`$DataInfo`) returns basic data information including the name of the Assemblage, sample size (`n`), total abundance (`N`), sampling fraction (`rho`), observed species richness (`S.obs`), sample coverage estimate of the reference sample with size n (`SC(n)`), sample coverage estimate of the extrapolated sample with size 2n (`SC(2n)`) as well as the first five species abundance frequency counts in the reference sample (`f1-f5`). If only data information is required, the simpler function `DataInfoIE()` (see later text) can be used to obtain the same output. More information about the maximum likelihood estimation (for any order q between 0 and 2) can be obtained by function `MLEAsyIE()`, which will be introduced later.

```{r eval=FALSE}
set.seed(2025)
data = list("Site_1" = as.numeric(rmultinom(1, 200, 1:30)), 
            "Site_2" = as.numeric(rmultinom(1, 300, 1:40)))

output_iNEXT <- iNEXTIE(data, rho = 0.3, q = c(0, 1, 2))
output_iNEXT$DataInfo
```

```{r echo=FALSE}
output_iNEXT[1]$DataInfo$`SC(n)`  <- round(output_iNEXT[1]$DataInfo$`SC(n)`, 3)
output_iNEXT[1]$DataInfo$`SC(2n)` <- round(output_iNEXT[1]$DataInfo$`SC(2n)`, 3)
output_iNEXT[1]
```

The second list of the output (`$iNextEst`) includes size- and coverage-based standardized diversity estimates and related statistics computed for 40 knots by default. There are two data frames (`$size_based` and `$coverage_based`). 

The first data frame (`$size_based`) includes the name of the assemblage (`Assemblage`), diversity order (`Order.q`), the target sample size (`m`), the `Method` (`Rarefaction`, `Observed`, or `Extrapolation`, depending on whether the size `m` is less than, equal to, or greater than the reference sample size), the diversity estimate of order q (`qIE`), the lower and upper confidence limits of diversity (`qIE.LCL` and `qIE.UCL`) conditioning on the sample size, and the corresponding sample coverage estimate (`SC`) along with the lower and upper confidence limits of sample coverage (`SC.LCL` and `SC.UCL`). These sample coverage estimates with confidence intervals are used for plotting the sample completeness curve. If the argument `nboot` is greater than zero, then a bootstrap method is applied to obtain the confidence intervals for the diversity and sample coverage estimates. Otherwise, all confidence intervals will not be computed. Here only the first six rows of the `$size_based` output are displayed:

```{r eval=FALSE}
output_iNEXT$iNextEst$size_based
```

```{r, echo=FALSE}
output_iNEXT$iNextEst$size_based$qIE     <- round(output_iNEXT$iNextEst$size_based$qIE, 3)
output_iNEXT$iNextEst$size_based$qIE.LCL <- round(output_iNEXT$iNextEst$size_based$qIE.LCL, 3)
output_iNEXT$iNextEst$size_based$qIE.UCL <- round(output_iNEXT$iNextEst$size_based$qIE.UCL, 3)
output_iNEXT$iNextEst$size_based$SC      <- round(output_iNEXT$iNextEst$size_based$SC, 3)
output_iNEXT$iNextEst$size_based$SC.LCL  <- round(output_iNEXT$iNextEst$size_based$SC.LCL, 3)
output_iNEXT$iNextEst$size_based$SC.UCL  <- round(output_iNEXT$iNextEst$size_based$SC.UCL, 3)
output_iNEXT$iNextEst$size_based[1:6,]
```

The second data frame (`$coverage_based`) includes the name of assemblage (`Assemblage`), the diversity order (`Order.q`), the target sample coverage value (`SC`), the corresponding sample size (`m`), the `Method` (`Rarefaction`, `Observed`, or `Extrapolation`, depending on whether the coverage `SC` is less than, equal to, or greater than the reference sample coverage), the diversity estimate of order q (`qIE`), the lower and upper confidence limits of diversity (`qIE.LCL` and `qIE.UCL`) conditioning on the target sample coverage value. Here only the first six rows of the `$coverage_based` output are displayed below: (Note for a fixed coverage value, the confidence interval in the `$coverage_based` table is wider than the corresponding interval in the `$size_based` table. This is because, for a given coverage value, the sample size needed to attain a fixed coverage value varies with bootstrap replication, leading to higher uncertainty on the resulting diversity estimate.)


```{r, eval=FALSE}
output_iNEXT$iNextEst$coverage_based
```

```{r, echo=FALSE}
output_iNEXT$iNextEst$coverage_based$qIE     <- round(output_iNEXT$iNextEst$coverage_based$qIE, 3)
output_iNEXT$iNextEst$coverage_based$qIE.LCL <- round(output_iNEXT$iNextEst$coverage_based$qIE.LCL, 3)
output_iNEXT$iNextEst$coverage_based$qIE.UCL <- round(output_iNEXT$iNextEst$coverage_based$qIE.UCL, 3)
output_iNEXT$iNextEst$coverage_based$SC      <- round(output_iNEXT$iNextEst$coverage_based$SC, 3)
output_iNEXT$iNextEst$coverage_based$m       <- round(output_iNEXT$iNextEst$coverage_based$m, 3)
output_iNEXT$iNextEst$coverage_based[1:6,]
```

The third list of the output (`$AsyEst`) includes the name of the Assemblage, diversity order (q = 0, 1, 2), the maximum likelihood estimation estimate (`IE_MLE`), asymptotic diversity estimate (`IE_asy`) and its estimated bootstrap standard error (`s.e.`) as well as the confidence intervals for asymptotic diversity (`qIE.LCL` and `qIE.UCL`). These statistics are computed only for q = 0, 1 and 2. More detailed information about maximum likelihood estimation and asymptotic diversity estimates for any order q between 0 and 2 can be obtained from function `MLEAsyIE()`. The output for `$AsyEst` is shown below:

```{r, eval=FALSE}
output_iNEXT$AsyEst
```

```{r, echo=FALSE}
tmp = output_iNEXT$AsyEst
tmp[,-(1:2)] = round(tmp[,-(1:2)], 3)
tmp
```

The `ggiNEXTIE` function can be used to make graphical displays for rarefaction and extrapolation sampling curves. An example for showing sample-size-based rarefaction/extrapolation curves (`type = 1`) is given below:

```{r}
# Sample-size-based R/E curves
ggiNEXTIE(output_iNEXT, type = 1)
```


The following commands return the sample completeness (sample coverage) curve (`type = 2`) in which different colors represent different assemblages. 

```{r fig.width = 5, fig.height = 4}
# Sample completeness curves for abundance data
ggiNEXTIE(output_iNEXT, type = 2)
```

The following commands return the coverage-based rarefaction/extrapolation sampling curves in which different color curves represent two assemblages within each diversity order.

```{r}
# Coverage-based R/E curves
ggiNEXTIE(output_iNEXT, type = 3)
```



## <span style="color:red;">FUNCTION DataInfoIE(): DATA INFORMATION</span> 

The function `DataInfoIE()` provides basic data information for the reference sample in each assemblage. The function `DataInfoIE()` with default arguments is shown below:
 
```{r eval=FALSE}
DataInfoIE(data, rho) 
```

All arguments in the above function are the same as those for the main function `iNEXTIE`. Running the `DataInfoIE()` function returns basic data information including sample size, total abundance, sampling fraction, observed species richness, sample coverage estimates of the reference sample (`SC(n)`), sample coverage estimate for twice the reference sample size (`SC(2n)`). We use data to demo the function.

### Basic data information for abundance data

```{r eval=FALSE}
set.seed(2025)
data = list("Site_1" = as.numeric(rmultinom(1, 200, 1:30)), 
            "Site_2" = as.numeric(rmultinom(1, 300, 1:40)))
DataInfoIE(data, rho = 0.3)
```

```{r echo=FALSE}
tmp <- DataInfoIE(data, rho = 0.3)
tmp$`SC(n)`  <- round(tmp$`SC(n)`, 3)
tmp$`SC(2n)` <- round(tmp$`SC(2n)`, 3)
tmp
```

Output description:

- `Assemblage` = assemblage name. 

- `n` = number of observed individuals in the reference sample (sample size).

- `N` = total abundance in the overall assemblage.

- `rho` = sampling fraction of the reference sample.

- `S.obs` = number of observed species in the reference sample.

- `SC(n)` = sample coverage estimate of the reference sample with size n.

- `SC(2n)` = sample coverage estimate of the reference sample with size 2n.

- `f1`-`f5` = the first five species abundance frequency counts in the reference sample.


## <span style="color:red;">FUNCTION estimateIE(): POINT ESTIMATION</span>

`estimateIE` is used to compute diversity estimates with q = 0, 1, 2 under any specified levels of sample size (when `base = "size"`) and sample coverage values (when `base = "coverage"`) for abundance data. When `base = "size"`, `level` can be specified with a particular vector of sample sizes (greater than 0); if `level = NULL`, this function computes the diversity estimates for the minimum sample size among all samples extrapolated to the double reference sizes or total individuals. When `base = "coverage"`, `level` can be specified with a particular vector of sample coverage values (between 0 and 1); if `level = NULL`, this function computes the diversity estimates for the minimum sample coverage among all samples extrapolated to the double reference sizes or total individuals. All arguments in the function are the same as those for the main function `iNEXTIE`.

```{r eval=FALSE}
estimateIE(data, rho, q = c(0, 1, 2), base = "coverage", level = NULL, nboot = 50, conf = 0.95) 
```

### Example: abundance data with two target coverage values (97% and 99%)

The following commands return the estimates with two specified levels of sample coverage (97% and 99%).  

```{r, eval=FALSE}
set.seed(2025)
data = list("Site_1" = as.numeric(rmultinom(1, 200, 1:30)), 
            "Site_2" = as.numeric(rmultinom(1, 300, 1:40)))

output_est_cov <- estimateIE(data, rho = 0.3, q = c(0, 1, 2), base = "coverage", level = c(0.97, 0.99))
output_est_cov
```

```{r, echo=FALSE}
output_est_cov <- estimateIE(data, rho = 0.3, q = c(0, 1, 2), base = "coverage", level = c(0.97, 0.99))

output_est_cov[,c('SC', 'm', 'qIE', 's.e.', 'qIE.LCL', 'qIE.UCL')] = 
  round(output_est_cov[,c('SC', 'm', 'qIE', 's.e.', 'qIE.LCL', 'qIE.UCL')], 3)

output_est_cov
```


### Example: abundance data with two target sample sizes (150 and 250)

The following commands return the estimates with two specified levels of sample sizes (150 and 250).

```{r, eval=FALSE}
set.seed(2025)
data = list("Site_1" = as.numeric(rmultinom(1, 200, 1:30)), 
            "Site_2" = as.numeric(rmultinom(1, 300, 1:40)))

output_est_size <- estimateIE(data, rho = 0.3, q = c(0, 1, 2), base = "size", level = c(150, 250))
output_est_size
```

```{r, echo=FALSE}
output_est_size <- estimateIE(data, rho = 0.3, q = c(0, 1, 2), base = "size", level = c(150, 250))

output_est_size[,c('m', 'SC', 'qIE', 's.e.', 'qIE.LCL', 'qIE.UCL')] = 
  round(output_est_size[,c('m', 'SC', 'qIE', 's.e.', 'qIE.LCL', 'qIE.UCL')], 3)

output_est_size
```


## <span style="color:red;">FUNCTION MLEAsyIE: MAXIMUM LIKELIHOOD ESTIMATION AND ASYMPTOTIC DIVERSITY PROFILES</span>

```{r, eval=FALSE}
MLEAsyIE(data, rho, q = seq(0, 2, 0.2), nboot = 50, conf = 0.95, method = c('Asymptotic', 'MLE'))
```

All arguments in the above function are the same as those for the main function `iNEXTIE` (except that the default of `q` here is `seq(0, 2, 0.2)`). The function `MLEAsyIE()` computes maximum likelihood estimation and asymptotic diversity of order q between 0 and 2 (in increments of 0.2) for diversity; these values with different order q can be used to depict a q-profile in the `ggMLEAsyIE` function. By default, both the maximum likelihood estimation and asymptotic diversity estimates will be computed.


## <span style="color:red;">FUNCTION ggMLEAsyIE(): GRAPHIC DISPLAYS OF DIVERSITY PROFILES</span>

```{r,eval=FALSE}
ggMLEAsyIE(output)
```

`ggMLEAsyIE` is a ggplot2 extension for an `MLEAsyIE` object to plot q-profile (which depicts the maximum likelihood estimation and asymptotic diversity estimate with respect to order q) for q between 0 and 2 (in increments of 0.2).

In the plot of profiles, only confidence intervals of the asymptotic diversity will be shown when both the maximum likelihood estimation and asymptotic diversity estimates are computed.


### Example: q-profiles for abundance data

The following commands returns the maximum likelihood estimation and asymptotic taxonomic diversity, along with its confidence interval for diversity order q between 0 to 2. Here only the first ten rows of the output are shown.

```{r, eval=FALSE}
set.seed(2025)
data = list("Site_1" = as.numeric(rmultinom(1, 200, 1:30)), 
            "Site_2" = as.numeric(rmultinom(1, 300, 1:40)))
output_MLEAsy <- MLEAsyIE(data, rho = 0.3)
output_MLEAsy
```

```{r, echo=FALSE, cache=TRUE}
output_MLEAsy <- MLEAsyIE(data, rho = 0.3)

tmp = output_MLEAsy
tmp[,(3:6)] <- round(tmp[,(3:6)], 3)
tmp[1:10,]
```

The following commands plot the corresponding q-profiles, along with its confidence interval for q between 0 to 2.

```{r}
ggMLEAsyIE(output_MLEAsy)
```




## License
The iNEXT.IE package is licensed under the GPLv3. To help refine `iNEXT.IE`, your comments or feedback would be welcome (please send them to Anne Chao or report an issue on the iNEXT.IE github [iNEXT.IE_github](https://github.com/KaiHsiangHu/iNEXT.IE)). 

## References


